---
title: "GS4PB ME Workflow: Multi-Environment Genomic Prediction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GS4PB ME Workflow: Multi-Environment Genomic Prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, comment = "#>")
```

# Overview

This vignette covers the **Multi-Environment (ME) genomic prediction workflow** in
GS4PB in detail.  It extends the brief ME section of the main pipeline vignette
with a focus on three advanced use cases that are common in HPC breeding programs:

| Step | Function | Purpose |
|------|----------|---------|
| ME Cross-Validation | `getME_CV()` | Estimate predictive ability across environments |
| Leave-One-Test-Out CV | `fitMEModels_LOFO_Pred(..., maskProp = 1)` | Evaluate prediction into completely new test environments |
| LOFO Prediction | `fitMEModels_LOFO_Pred(..., maskProp < 1)` | Predict partially-observed test environments |

All code chunks use `eval = FALSE` because the pipeline requires real genotype and
phenotype files.  Replace every `<...>` placeholder before running.

The workflow is designed to run on an **HPC cluster** using `foreach %dopar%` to
parallelize prediction across multiple factor levels (e.g., year × test
combinations).

---

# 1 — HPC Setup

When running on a cluster that uses environment modules, load the required system
libraries before starting R.  The exact module names vary by cluster; the example
below reflects a typical SLURM / Rocky8 setup.

```{r hpc-modules, engine="bash"}
## HPC module loads (run in your job script or interactive session)
## Adjust module names to match your cluster's module system
module load udunits
module load gdal
module load proj
module load netcdf
module load geos
module load R/4.4.2-openblas-rocky8
R
```

```{r libs}
## Set Java heap size BEFORE loading rJava — cannot be changed after JVM starts
options(java.parameters = "-Xmx30g")

library(rJava)
library(rTASSEL)
library(GS4PB)
library(dplyr)
library(foreach)
library(doParallel)

## Start the TASSEL logger (suppresses verbose Java output to console)
rTASSEL::startLogger()
```

---

# 2 — Define Paths and Parameters

```{r inputs}
## ── File paths ──────────────────────────────────────────────────────────────
vcf_path      <- "<path/to/genotype.vcf>"
me_pheno_path <- "<path/to/me_phenotype.csv>"
out_dir       <- "<path/to/output_directory>"

## ── Genotype QC parameters ──────────────────────────────────────────────────
MAF_thresh       <- 0.02    # minor allele frequency threshold
site_call_rate   <- 0.80    # minimum proportion of non-missing calls per site
taxa_min_called  <- 0.90    # minimum proportion of non-missing calls per taxon

## ── Trait and column name mapping ───────────────────────────────────────────
## Traits to analyse — one or more column names in the phenotype file
trait_me    <- c("Yield")

## Column in the phenotype file that identifies each genotype/strain
strain_col  <- "Strain"

## Column used as the environmental grouping variable (e.g., "Env", "Loc")
env_col     <- "Env"

## Location and year columns — set to NULL if not present
loc_col     <- "Loc"    # or NULL
yr_col      <- "Year"   # or NULL

## Column containing the unique observation ID (strain × environment key)
## Used internally for data alignment; e.g., "UniqID", "StrLocTst"
id_col_me   <- "UniqID"

## Fixed effect to include in ME models — typically the environment variable
fix_me      <- "Env"

## ── ME model parameters ─────────────────────────────────────────────────────
## Kernel method: "Linear" (GB) builds a linear genomic kernel;
##                "Gaussian" (GK) builds a Gaussian kernel — often more flexible
k_method    <- "GK"

## Scope filters: "All" uses all locations / years; or supply a character vector
loc_me      <- "All"
yr_me       <- "All"

## ── Bayesian MCMC parameters (BGGE backend) ─────────────────────────────────
bys_iter   <- 10000
bys_burnin <- 2000
bys_thin   <- 5

## ── Cross-validation parameters ─────────────────────────────────────────────
k_folds <- 5    # number of CV folds
n_iter  <- 5    # number of CV replicates
cv_met  <- "CV0"  # "CV1" (novel genotypes), "CV2" (novel GxE cells),
                  # "CV0" (novel environments), "CV00" (novel geno x novel env)

## ── LOFO / LOTCV factor setup ────────────────────────────────────────────────
## Column whose levels are used as the leave-one-out factor
## A common choice is a composite year × test label, e.g., "2025-T1"
factor_col <- "YrTst"

## Number of parallel workers for the foreach loop
n_workers  <- 10
```

---

# 3 — Genotype Processing

Genotype processing is shared across all GS4PB workflows.  See the main vignette
(`vignette("GS4PB_Pipeline")`) for a full description.  The steps below are
reproduced for self-contained reference.

## 3a — Load VCF

`getTasObj()` reads the VCF into JVM memory via rTASSEL.
`getGenoTas_to_DF()` converts it to an R data frame coded in 0/1/2 format.

```{r load-vcf}
tas     <- getTasObj(vcf_path)
geno_df <- getGenoTas_to_DF(tas)
cat(getGenoQCStats(geno_df))
## Reports: number of genotypes, number of markers, % missing
```

## 3b — Filter Markers (Filt1)

Remove SNPs below the MAF threshold or with a low site call rate.
`siteMinCnt` expects an **integer count**, not a proportion — multiply by the
number of markers and round to avoid rTASSEL segmentation faults.

```{r filt1}
n_markers <- ncol(geno_df) - 5L   # subtract non-marker columns

tas_filt1  <- getFilteredSitesGenoData(tas,
                siteMinCnt = round(site_call_rate * n_markers),
                MAF        = MAF_thresh)
geno_filt1 <- getGenoTas_to_DF(tas_filt1)
cat(getGenoQCStatsFilt1(geno_df, geno_filt1,
                        missSitesTH = 1 - site_call_rate,
                        MAFTH       = MAF_thresh))
```

## 3c — Filter Samples (Filt2)

Remove individuals with high missingness across retained markers.

```{r filt2}
tas_filt2  <- getFilteredTaxaGenoData(tas_filt1,
                MinNotMissing = taxa_min_called)
geno_filt2 <- getGenoTas_to_DF(tas_filt2)
cat(getGenoQCStatsFilt2(geno_filt1, geno_filt2,
                        missIndTH = 1 - taxa_min_called))
```

## 3d — Imputation

Choose **one** method.  LD-KNNi generally performs better than numeric
imputation when the dataset is large and has low to moderate missingness.

```{r impute}
## Option A: Numeric (mean) imputation — fast, suitable for low missingness
# tas_imp  <- getImputedData_Num(tas_filt2, nN = 5, Dist = "Euclidean")
# geno_imp <- getGenoTas_to_DF(tas_imp)

## Option B: LD-KNNi imputation via rTASSEL — more accurate, slower
tas_imp  <- getImputedData_LDKNNI(tas_filt2, l = 30, k = 10)
geno_imp <- getGenoTas_to_DF(tas_imp)

cat(getGenoQCStats(geno_imp))
## Should report 0% missing after imputation
```

---

# 4 — ME Phenotype Setup

## 4a — Load and Clean the Phenotype File

The ME phenotype CSV must have **one row per observation** (strain × environment).
Required columns: strain ID, environment, trait value(s).  Optional: Loc, Year.

```{r me-pheno-load}
pheno_me_raw <- read.csv(me_pheno_path, stringsAsFactors = FALSE)

## Strip any prefix added by export tools (e.g., "predicted.value." from ASReml)
colnames(pheno_me_raw) <- gsub("predicted\\.value\\.", "", colnames(pheno_me_raw))

## Remove rows with missing Test labels or any unwanted test identifiers
## Adjust the pattern to match entries you wish to exclude
exclude_idx <- grep("^NA$|^$", pheno_me_raw$Test)
if (length(exclude_idx) > 0) {
  pheno_me_raw <- pheno_me_raw[-exclude_idx, ]
}
```

## 4b — Create Composite Factor Column

A composite year × test label is useful for leave-one-test-out analyses.
This column (`factor_col`) becomes the grouping variable for the LOFO loop.

```{r factor-col}
## Paste Year and Test (or any two columns) to create a unique factor level
## per year × test combination — e.g., "2025-T1", "2024-T2"
pheno_me_raw[[factor_col]] <- paste(pheno_me_raw$Year,
                                    pheno_me_raw$Test,
                                    sep = "-")
```

## 4c — Identify ID and Trait Columns

```{r id-cols}
## All columns that are NOT trait values are treated as ID / metadata columns
id_cols_me <- colnames(pheno_me_raw)[!colnames(pheno_me_raw) %in% trait_me]
```

## 4d — Standardise Column Names

`stdizePhVarNames()` remaps user-supplied column names to the internal names
(`env`, `gid`, `uniqID`) expected by all downstream ME functions.

```{r standardise}
std_out     <- stdizePhVarNames(pheno_me_raw,
                 IDColsME    = id_cols_me,
                 StrainME    = strain_col,
                 EnvVarME    = env_col,
                 LocMEColVar = loc_col,
                 YrMEColVar  = yr_col)

pheno_me_mod <- std_out[[1]]   # standardised phenotype data frame
id_cols_mod  <- std_out[[2]]   # updated ID column vector (with renamed entries)
env_var_mod  <- std_out[[3]]   # standardised environment variable name
```

## 4e — Format into Per-Trait Data Frames

`getPhenoMEData()` splits the phenotype table into a list of per-trait data
frames.  Each element is a long-format table with columns `env`, `gid`,
`value`, and the metadata columns from `id_cols_mod`.

```{r me-pheno-format}
pheno_me_fmt <- getPhenoMEData(
  PhenoME    = pheno_me_mod,
  TraitME    = trait_me,
  nSelTraits = length(trait_me),
  IDColsME   = id_cols_mod,
  StrainME   = strain_col
)
```

## 4f — Merge Genotype and Phenotype

`getMergedDataME()` aligns strain IDs across the genotype matrix and each
per-trait phenotype table.  Only strains present in both are retained.

```{r me-merge}
merged_me <- getMergedDataME(phData    = pheno_me_fmt,
                              genoImp   = geno_imp,
                              TargetIDs = NULL)

DT_list   <- merged_me[[1]]   # list of phenotype data frames, one per trait
geno_list <- merged_me[[2]]   # list of genotype matrices, one per trait

cat("Genotypes in merged data:", length(unique(DT_list[[1]]$Strain)), "\n")
```

---

# 5 — ME Cross-Validation

`getME_CV()` fits multi-environment BGGE models under a chosen CV scheme and
returns fold-level predictions and variance components.

**CV schemes:**

| `CVMet` | Scenario tested |
|---------|----------------|
| `"CV1"` | Novel genotypes in observed environments (hardest) |
| `"CV2"` | Observed genotypes in partially-observed environments |
| `"CV0"` | Novel (held-out) environments |
| `"CV00"` | Novel genotypes × novel environments |
| `"CV_LOFO"` | Leave-one-factor-level-out (e.g., one location) |

```{r me-cv}
## Register parallel backend for the within-CV fold loop
registerDoParallel(n_workers)

cv_me <- getME_CV(
  DT_1_Filt_List = DT_list,
  genoDat_List   = geno_list,
  traits         = trait_me,
  KG             = NULL,           # NULL = G matrix estimated internally
  KE             = NULL,           # NULL = no environmental kernel
  CVMet          = cv_met,
  factVar        = env_var_mod,
  K              = k_folds,
  NIter          = n_iter,
  KMethod        = k_method,       # "Linear" (GB) or "Gaussian" (GK)
  FitEnvModels   = FALSE,          # TRUE only when KE is provided
  fixedME        = fix_me,
  envVar         = env_var_mod,
  IDColsME       = id_cols_mod,
  IDColME        = id_col_me,
  LocME          = loc_me,
  YrME           = yr_me,
  b_iter         = bys_iter,
  b_burnin       = bys_burnin,
  b_thin         = bys_thin
)

## cv_me$results[[trait_idx]] contains per-rep, per-fold predictions
## and variance components for models MM, MDs, MDe (and SM for single-env folds)
cv_me_fit <- cv_me$results[[1]]
```

---

# 6 — Leave-One-Test-Out Cross-Validation (LOTCV)

LOTCV asks: *"How well does the model predict an entirely new test environment,
given data from all other environments?"*

`fitMEModels_LOFO_Pred()` with `maskProp = 1` fully removes all observations
from `maskFactrLev` before fitting, then predicts them.  Running one call per
target factor level (parallelised over test sets) produces a complete LOTCV
assessment.

## 6a — Identify Target Factor Levels

```{r lotcv-setup}
## All unique levels of the composite factor column (e.g., "2025-T1", "2025-T2")
all_factor_levels    <- unique(DT_list[[1]][[factor_col]])

## Subset to the levels you want to predict — here, those from a specific year
target_factor_levels <- grep("2025", all_factor_levels, value = TRUE)
n_targets            <- length(target_factor_levels)

cat("Factor levels to predict:", n_targets, "\n")
print(target_factor_levels)
```

## 6b — Run LOTCV in Parallel

Each parallel task masks one factor level completely (`maskProp = 1`), fits
MM / MDs / MDe models on the remaining data, and returns predictions for the
held-out level.

`withCallingHandlers` captures convergence warnings without aborting the job;
`tryCatch` catches hard errors and returns `NULL` for that level so the loop
continues.

```{r lotcv-run}
registerDoParallel(n_workers)

lotcv_results <- foreach(
  i = seq_len(n_targets),
  .packages = c("GS4PB")
) %dopar% {
  tryCatch({
    withCallingHandlers({

      GS4PB::fitMEModels_LOFO_Pred(
        DT_1_Filt_List = DT_list,
        genoDat_List   = geno_list,
        traits         = trait_me,
        KG             = NULL,
        KE             = NULL,
        factrVar       = factor_col,
        maskFactrLev   = target_factor_levels[i],   # mask this level entirely
        maskProp       = 1,                          # 1 = full leave-one-out
        nSampleReps    = 1,
        KMethod        = k_method,
        fixedME        = fix_me,
        envVar         = env_var_mod,
        IDColsME       = id_cols_mod,
        IDColME        = id_col_me,
        LocME          = loc_me,
        YrME           = yr_me,
        b_iter         = bys_iter,
        b_burnin       = bys_burnin,
        b_thin         = bys_thin
      )

    }, warning = function(w) {
      ## Capture but do not abort on convergence warnings
      msg <- conditionMessage(w)
      message("Warning [", target_factor_levels[i], "]: ", msg)
      invokeRestart("muffleWarning")
    })
  }, error = function(e) {
    message("Error [", target_factor_levels[i], "]: ", e$message)
    NULL   # return NULL for this level; downstream code handles it
  })
}

names(lotcv_results) <- target_factor_levels
```

## 6c — Summarise LOTCV Predictive Ability

`getCombinedTab()` merges predictions from all three models (MM, MDs, MDe) and
observed values into a single wide data frame per factor level.

```{r lotcv-summary}
n_trait <- 1   # index of the trait within trait_me (1 = first trait)

lotcv_env_cor <- do.call(rbind, lapply(seq_len(n_targets), function(i) {

  res <- lotcv_results[[i]]
  if (is.null(res)) return(NULL)   # skip levels that errored

  combined <- getCombinedTab(
    outputListME = res,
    TraitME      = trait_me,
    IDColsME     = id_cols_mod,
    IDColME      = id_col_me,
    envVar       = env_var_mod,
    fitEnvCov    = FALSE,
    reaction     = FALSE
  )

  ## Column names follow the pattern "<Trait>Pred-<Model>" and "<Trait>Obs-<Model>"
  trait  <- trait_me[n_trait]
  cor_tab <- combined %>%
    group_by(Env) %>%
    summarise(
      Cor_MM  = cor(get(paste0(trait, "Pred-MM")),
                    get(paste0(trait, "Obs-MM")),  use = "pairwise.complete.obs"),
      Cor_MDs = cor(get(paste0(trait, "Pred-MDs")),
                    get(paste0(trait, "Obs-MDs")), use = "pairwise.complete.obs"),
      Cor_MDe = cor(get(paste0(trait, "Pred-MDe")),
                    get(paste0(trait, "Obs-MDe")), use = "pairwise.complete.obs"),
      .groups = "drop"
    )

  cor_tab$FactorLevel <- target_factor_levels[i]
  cor_tab
}))

print(lotcv_env_cor)
```

---

# 7 — LOFO Prediction with Partial Masking

LOFO prediction (`maskProp < 1`) is appropriate when **some** observations from
the target environment are available for training — for example, early-season
data exist but the full harvest is not yet recorded.  The partial-masking regime
samples `maskProp` of observations as test, fits on the rest, and repeats
`nSampleReps` times to average out sampling variance.

## 7a — Run LOFO Prediction in Parallel

```{r lofo-run}
n_sample_reps <- 10   # number of partial-mask replicates per factor level
mask_prop     <- 0.5  # proportion of observations masked as test set

registerDoParallel(n_workers)

lofo_pred_results <- foreach(
  i = seq_len(n_targets),
  .packages = c("GS4PB")
) %dopar% {
  tryCatch({
    withCallingHandlers({

      GS4PB::fitMEModels_LOFO_Pred(
        DT_1_Filt_List = DT_list,
        genoDat_List   = geno_list,
        traits         = trait_me,
        KG             = NULL,
        KE             = NULL,
        factrVar       = factor_col,
        maskFactrLev   = target_factor_levels[i],
        maskProp       = mask_prop,        # partial masking
        nSampleReps    = n_sample_reps,    # repeat n times, average predictions
        KMethod        = k_method,
        fixedME        = fix_me,
        envVar         = env_var_mod,
        IDColsME       = id_cols_mod,
        IDColME        = id_col_me,
        LocME          = loc_me,
        YrME           = yr_me,
        b_iter         = bys_iter,
        b_burnin       = bys_burnin,
        b_thin         = bys_thin
      )

    }, warning = function(w) {
      msg <- conditionMessage(w)
      message("Warning [", target_factor_levels[i], "]: ", msg)
      invokeRestart("muffleWarning")
    })
  }, error = function(e) {
    message("Error [", target_factor_levels[i], "]: ", e$message)
    NULL
  })
}

names(lofo_pred_results) <- target_factor_levels
```

## 7b — Average Predictions Across Sampling Replicates

For each factor level, average the replicated predictions from each model across
`nSampleReps` iterations, then join MM, MDs, MDe into a single table.

```{r lofo-avg}
n_trait    <- 1        # trait index
trait_name <- trait_me[n_trait]

avg_pred_list   <- vector("list", n_targets)
env_cor_list    <- vector("list", n_targets)
best_model_list <- vector("list", n_targets)

for (i in seq_len(n_targets)) {

  res <- lofo_pred_results[[i]]
  if (is.null(res)) next

  ## Average each model's predictions across sampling replicates
  ## res$preds[[trait_idx]]$MM  is a data frame with cols: Strain, Env, Pred, Obs, Rep
  avg_MM  <- res$preds[[n_trait]]$MM  %>%
    group_by(Strain, Env) %>%
    summarise(Pred.MM  = mean(Pred, na.rm = TRUE),
              Obs      = mean(Obs,  na.rm = TRUE), .groups = "drop")

  avg_MDs <- res$preds[[n_trait]]$MDs %>%
    group_by(Strain, Env) %>%
    summarise(Pred.MDs = mean(Pred, na.rm = TRUE), .groups = "drop")

  avg_MDe <- res$preds[[n_trait]]$MDe %>%
    group_by(Strain, Env) %>%
    summarise(Pred.MDe = mean(Pred, na.rm = TRUE), .groups = "drop")

  ## Join all three model predictions into one table
  avg_tab <- avg_MM %>%
    inner_join(avg_MDs, by = c("Strain", "Env")) %>%
    inner_join(avg_MDe, by = c("Strain", "Env"))

  ## Rename to interpretable model names
  ## MM  = G + E (main effects only)
  ## MDs = G + E + GxE (single variance component for GxE)
  ## MDe = G + E + GxEi (environment-specific GxE variance)
  colnames(avg_tab) <- c("Strain", "Env", "G+E", "Obs", "G+E+GxE", "G+E+GxEi")

  avg_tab$FactorLevel <- target_factor_levels[i]
  avg_pred_list[[i]]  <- avg_tab

  ## Compute per-environment correlations for each model
  cor_tab <- avg_tab %>%
    group_by(FactorLevel, Env) %>%
    summarise(
      Cor_MM  = cor(`G+E`,       Obs, use = "pairwise.complete.obs"),
      Cor_MDs = cor(`G+E+GxE`,  Obs, use = "pairwise.complete.obs"),
      Cor_MDe = cor(`G+E+GxEi`, Obs, use = "pairwise.complete.obs"),
      .groups = "drop"
    )

  colnames(cor_tab) <- c("FactorLevel", "Env", "G+E", "G+E+GxE", "G+E+GxEi")
  env_cor_list[[i]] <- cor_tab

  ## Identify the best-performing model per environment
  if (nrow(cor_tab) > 1) {
    best_mod <- apply(cor_tab, 1,
                      function(x) names(x[3:5])[which.max(as.numeric(x[3:5]))])
  } else {
    best_mod <- names(cor_tab[3:5])[which.max(cor_tab[3:5])]
  }
  names(best_mod)        <- cor_tab$Env
  best_model_list[[i]]   <- best_mod
}
```

## 7c — Build Final Prediction Table Using Best Model per Environment

For each environment, select the predictions from whichever model had the
highest correlation.  If no valid correlation exists (e.g., all-NA environment),
fall back to the average of MM and MDs.

```{r lofo-final-table}
final_pred_table <- c()

for (i in seq_along(avg_pred_list)) {

  avg_tab <- avg_pred_list[[i]]
  if (is.null(avg_tab)) next

  envs    <- as.character(unique(avg_tab$Env))
  out     <- c()

  for (env in envs) {

    best_mod_env <- as.character(best_model_list[[i]][env])
    x_env        <- avg_tab[avg_tab$Env == env, ]

    if (best_mod_env != "character(0)" && best_mod_env %in% colnames(x_env)) {
      ## Use the single best model prediction for this environment
      out_env              <- x_env[, c("Strain", "Obs", best_mod_env,
                                        "FactorLevel", "Env")]
      colnames(out_env)    <- c("Strain", "Obs", "Pred", "FactorLevel", "Env")
    } else {
      ## Fallback: average MM and MDs predictions
      mod_cols  <- which(colnames(x_env) %in% c("G+E", "G+E+GxE"))
      avg_preds <- apply(x_env[, mod_cols], 1, function(r) mean(as.numeric(r)))
      out_env   <- cbind(x_env[, c("Strain", "Obs")],
                         Pred        = avg_preds,
                         FactorLevel = x_env$FactorLevel,
                         Env         = x_env$Env)
    }

    out <- rbind(out, out_env)
  }

  final_pred_table <- rbind(final_pred_table, out)
}

## Summary: best-model correlation per environment
final_env_cor <- final_pred_table %>%
  group_by(Env) %>%
  summarise(Cor_BestModel = cor(Obs, Pred, use = "pairwise.complete.obs"),
            .groups = "drop")

print(final_env_cor)
```

## 7d — Write Outputs

```{r lofo-write}
write.csv(final_pred_table,
          file      = file.path(out_dir, paste0(trait_name, "_LOFO_Predictions.csv")),
          quote     = FALSE,
          row.names = FALSE)

write.csv(final_env_cor,
          file      = file.path(out_dir, paste0(trait_name, "_LOFO_EnvCor.csv")),
          quote     = FALSE,
          row.names = FALSE)

cat("Results written to:", out_dir, "\n")
```

---

# 8 — Multi-Trait Extension

The ME workflow can be repeated for additional traits.  Traits may differ in how
the environment is defined — for example, a maturity rating may be grouped by
**location** rather than by environment × year.  Simply re-run Sections 4–7
with a new set of column-name mappings.

```{r multi-trait}
## ── Second trait example: trait grouped by location ─────────────────────────

trait_me_2   <- c("Maturity")       # trait column in the phenotype file
env_col_2    <- "Loc"               # environment grouping for this trait
loc_col_2    <- "Loc"
fix_me_2     <- "Env"

## Exclude this trait's column from the ID column list
id_cols_me_2 <- colnames(pheno_me_raw)[!colnames(pheno_me_raw) %in% trait_me_2]
## Also exclude columns belonging to other traits to keep ID columns clean
id_cols_me_2 <- grep("Maturity", id_cols_me_2, invert = TRUE, value = TRUE)

std_out_2    <- stdizePhVarNames(pheno_me_raw,
                  IDColsME    = id_cols_me_2,
                  StrainME    = strain_col,
                  EnvVarME    = env_col_2,
                  LocMEColVar = loc_col_2,
                  YrMEColVar  = yr_col)

pheno_me_mod_2 <- std_out_2[[1]]
id_cols_mod_2  <- std_out_2[[2]]
env_var_mod_2  <- std_out_2[[3]]

pheno_me_fmt_2 <- getPhenoMEData(
  PhenoME    = pheno_me_mod_2,
  TraitME    = trait_me_2,
  nSelTraits = length(trait_me_2),
  IDColsME   = id_cols_mod_2,
  StrainME   = strain_col
)

merged_me_2 <- getMergedDataME(phData    = pheno_me_fmt_2,
                                genoImp   = geno_imp,
                                TargetIDs = NULL)
DT_list_2   <- merged_me_2[[1]]
geno_list_2 <- merged_me_2[[2]]

## Re-run Sections 6–7 substituting:
##   DT_list   → DT_list_2
##   geno_list → geno_list_2
##   trait_me  → trait_me_2
##   id_cols_mod  → id_cols_mod_2
##   env_var_mod  → env_var_mod_2
##   fix_me       → fix_me_2
## All other parameters (k_method, bys_iter, etc.) remain the same.
```

---

# 9 — Tips for HPC Runs

**Memory:**  Set `options(java.parameters = "-Xmx30g")` *before* `library(rJava)`.
Changing it afterwards has no effect.

**Parallelism:**  `registerDoParallel(n_workers)` on a SLURM cluster should match
the `--cpus-per-task` value in your job script to avoid over-subscribing cores.

**BGGE convergence:**  Increase `bys_iter` and decrease `bys_thin` if you see
repeated convergence warnings.  A minimum of `iter = 10000, burnin = 2000,
thin = 5` is recommended for ME models.

**Marker matrix:**  Pass `KG = G_tuned` (pre-computed G matrix) to
`getME_CV()` and `fitMEModels_LOFO_Pred()` when the same genotype set is
analysed across multiple traits.  This avoids redundant G matrix computation
inside each parallel worker.

**Null results:**  If `lofo_pred_results[[i]]` is `NULL`, the worker hit an
error (printed to stderr).  Check that all levels in `target_factor_levels` are
present in `DT_list[[1]][[factor_col]]` and that each level has at least two
environments in the training set.
