---
title: "GS4PB Pipeline: From VCF to GEBV Predictions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GS4PB Pipeline: From VCF to GEBV Predictions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, comment = "#>")
```

# Overview

**GS4PB** provides an end-to-end genomic selection pipeline for plant breeding.
This vignette walks through the complete workflow — from loading a VCF file to
producing GEBV predictions — covering three scenarios:

| Workflow | Description |
|----------|-------------|
| **SE:ST** | Single-Environment, Single-Trait |
| **SE:MT** | Single-Environment, Multi-Trait |
| **ME**    | Multi-Environment |

All code chunks use `eval = FALSE` because the pipeline requires real genotype
and phenotype files. Copy, fill in the `<...>` placeholders, and run interactively.

A copy of the raw R script is also installed with the package:

```{r script-path}
system.file("pipeline", "GS4PB_Pipeline.R", package = "GS4PB")
```

---

# Prerequisites

```{r prereqs}
# install.packages("path/to/GS4PB_0.0.0.9000.tar.gz",
#                  repos = NULL, type = "source")

library(GS4PB)
rTASSEL::initializeTASSEL()   # start the JVM before any rTASSEL calls
```

## Define Paths and Parameters

Replace every `<...>` with real paths and column names before running.

```{r inputs}
vcf_path       <- "<path/to/genotype.vcf>"
se_pheno_path  <- "<path/to/se_phenotype.csv>"
me_pheno_path  <- "<path/to/me_phenotype.csv>"
target_id_path <- "<path/to/target_ids.txt>"   # optional
out_dir        <- "<path/to/output_directory>"

## SE trait names
trait_se   <- "Yield"
traits_mt  <- c("Yield", "Height", "Maturity")

## ME phenotype column names (standardised internally)
strain_col <- "Strain"
env_col    <- "Env"
loc_col    <- "Location"   # or NULL
yr_col     <- "Year"       # or NULL
trait_me   <- c("Yield")

## VCF filter parameters
MAF_thresh       <- 0.05
site_min_cnt     <- 0.80
taxa_min_nr_miss <- 0.80

## CV parameters
k_folds <- 5
n_iter  <- 5
cv_met  <- "CV1"   # "CV1" = by observation; "CV2" = by factor level
```

---

# Section 1 — Genotypic Data Processing

This section is shared by all three workflows.

## 1a — Load VCF

`rTASSEL` reads the VCF into JVM memory; `getGenoTas_to_DF()` converts it to an
R tibble with columns `SNPID`, `Chrom`, `Pos`, `REF`, `ALT`, and one column per
sample (coded 0/1/2).

```{r load-vcf}
tas     <- getTasObj(vcf_path)
geno_df <- getGenoTas_to_DF(tas)
cat(getGenoQCStats(geno_df))
```

## 1b — Filter Markers (Filt1)

Remove SNPs below the MAF threshold or with low call rate.

```{r filt1}
tas_filt1  <- getFilteredSitesGenoData(tas,
                siteMinCnt = site_min_cnt,
                MAF        = MAF_thresh)
geno_filt1 <- getGenoTas_to_DF(tas_filt1)
cat(getGenoQCStatsFilt1(geno_df, geno_filt1,
                        missSitesTH = 1 - site_min_cnt,
                        MAFTH       = MAF_thresh))
```

## 1c — Filter Samples (Filt2)

Remove individuals with high missingness.

```{r filt2}
tas_filt2  <- getFilteredTaxaGenoData(tas_filt1,
                MinNotMissing = taxa_min_nr_miss)
geno_filt2 <- getGenoTas_to_DF(tas_filt2)
cat(getGenoQCStatsFilt2(geno_filt1, geno_filt2,
                        missIndTH = 1 - taxa_min_nr_miss))
```

## 1d — Imputation

Choose **one** method; comment out the others.

```{r impute}
## Option A: Numeric (mean) imputation via rTASSEL  [DEFAULT]
tas_imp  <- getImputedData_Num(tas_filt2, nN = 5, Dist = "Euclidean")
geno_imp <- getGenoTas_to_DF(tas_imp)

## Option B: LD-KNNi imputation via rTASSEL
# tas_imp  <- getImputedData_LDKNNI(tas_filt2, l = 50, k = 5)
# geno_imp <- getGenoTas_to_DF(tas_imp)

## Option C: AlphaPlantImpute (external tool required)
# getGenoData_API(tas_filt2)              # writes current_GenoTable.genotypes
# # ... run AlphaPlantImpute externally ...
# geno_imp <- getImpGenoData_API(vcfIDTab) # reads back imputed_out.genotypes
```

## 1e — (Optional) Genomic Kinship / G Matrix

Compute the VanRaden genomic relationship matrix.  The tuned `G_tuned` can be
passed as a custom kernel to SE:MT and ME models.

```{r gmat}
qc_result <- getQCFilteredM(geno_imp, maf = MAF_thresh)
M_clean   <- qc_result$M.clean

gmat_out  <- getGmat(M_clean, method = "VanRaden")
G         <- gmat_out$G

## Detect and remove duplicate individuals
diag_out  <- getKinshipDiag(G, duplicate.thr = 0.98)
cat(diag_out$msg)

if (!is.null(diag_out$list.duplicate) &&
    nrow(diag_out$list.duplicate) > 0) {
  dedup_out <- removeDuplicatesFromG(G, diag_out$list.duplicate)
  G         <- dedup_out$G_clean
  cat(dedup_out$msg)
}

## Tune G for positive-definiteness
G_tuned    <- getTunedG(G, method = "blend", pblend = 0.05)$Gb

## PCA for visualisation
pca_result <- getKinshipPCA(G_tuned, ncp = 20)
# plot(pca_result$ind$coord[, 1], pca_result$ind$coord[, 2])
```

---

# Section 2 — SE Workflow: Single-Environment

## 2a — Load SE Phenotype

The CSV must have one row per individual, with a strain ID column and one or more
trait columns.

```{r se-pheno}
pheno_se   <- read.csv(se_pheno_path, stringsAsFactors = FALSE)
target_ids <- NULL   # or readLines(target_id_path)
```

## 2b — Merge Genotype + SE Phenotype

`getMergedData()` aligns strain IDs across genotype and phenotype data frames and
returns a list of train and test (target) subsets.

```{r se-merge}
merged    <- getMergedData(gt2d     = geno_imp,
                           Pheno    = pheno_se,
                           testIDs  = target_ids)
# merged[[1]] = training data frame
# merged[[2]] = test (target) data frame; NULL if target_ids = NULL
```

## 2c — Process Data

Remove rows where the focal trait is `NA`.

```{r se-process}
proc_data <- getProcessedData(merged, trait = trait_se)
```

## 2d — Sample Candidate Set

```{r se-predict-data}
n_cand    <- nrow(proc_data[[1]])   # use all training individuals
pred_data <- getPredictionData(proc_data, noCandidates = n_cand)
```

## 2e — (Optional) Training Set Optimization via Genetic Algorithm

Uncomment to select an optimal training subset with STPGA.

```{r opt-ts}
# Single-objective (errorstat = 1 value) — uses GenAlgForSubsetSelection
# ga_params <- list(
#   errorstat    = "MSE", InitPop = NULL, npop = 50, nelite = 5,
#   mutprob = 0.01, mutintensity = 1, niterations = 100,
#   minitbefstop = 10, tabu = FALSE, tabumemsize = 5,
#   plotiters = FALSE, lambda = 0.5, mc.cores = 2
# )
#
# Multi-objective (errorstat = vector, length > 1) — uses GenAlgForSubsetSelectionMO
# ga_params <- list(
#   errorstat           = c("PEVmean", "meanCD"),
#   selectionstatstypes = c("min", "min"),   # "min" or "max" per stat
#   plotdirections      = c("min", "min"),
#   InitPop = NULL, npop = 100, mutprob = 0.8, mutintensity = 1,
#   niterations = 500, plotiters = FALSE, lambda = 1e-6, mc.cores = 2
# )
# n_train_select <- 80
#
# opt_ts  <- getOptimalTS(pred_data, trait = trait_se, nTraits = 1,
#                         noCandidates = n_cand,
#                         nTrainToSelect = n_train_select,
#                         GAParameters = ga_params)
# rand_ts <- getRandomTS(pred_data, trait = trait_se, nTraits = 1,
#                        noCandidates = n_cand,
#                        nTrainToSelect = n_train_select)
# ts_cmp  <- getTSComparisons(pred_data,
#                              Train_STPGA  = opt_ts,
#                              Train_Random = rand_ts,
#                              trait = trait_se, nTraits = 1,
#                              testIds = rownames(pred_data[[2]]))
# print(ts_cmp)
# opt_ids <- opt_ts[[2]]   # selected training strain IDs
opt_ids <- NULL            # NULL = use all training data
```

---

## SE:ST — Single-Trait

### 2f — SE:ST Cross-Validation

5-fold CV using **bWGR** (emRR, emBB, emBL).  Returns prediction ability (PA)
and RMSE per model.

```{r se-st-cv}
cv_st <- getemCVR(pred_data,
                  trait  = trait_se,
                  nTraits = 1,
                  k      = k_folds,
                  nIter  = n_iter)
print(cv_st)
```

### 2g — SE:ST GEBV Prediction

Trains on the training set (or `opt_ids` subset) and ranks all individuals.

```{r se-st-gebv}
gebv_st <- getRankedPredictedValues(
  pred_data,
  nTraits  = 1,
  trait    = trait_se,
  GPModel  = "rrBLUP (rrBLUP)",  # "BayesB (bWGR)", "emRR (bWGR)", etc.
  optTS    = opt_ids
)
# gebv_st[[1]] = ranked training predictions (LineID, Predicted_Value)
# gebv_st[[2]] = ranked target predictions   (LineID, Predicted_Value, Reliability)

writeCVOutTable(as.data.frame(cv_st), type = "ST", outDirPath = out_dir)
writeGPOutTable(gebv_st[[1]],         type = "ST", outDirPath = out_dir)
```

---

## SE:MT — Multi-Trait

### 2h — SE:MT Cross-Validation

Bayesian multi-trait MCMC via **BGLR**.  Passing `A_ext = G_tuned` uses the
genomic kinship matrix as the relationship kernel; set `A_ext = NULL` to
estimate the A matrix internally with `rrBLUP::A.mat()`.

```{r se-mt-cv}
cv_mt <- getMTCVR(
  pred_data,
  trait    = traits_mt,
  nTraits  = length(traits_mt),
  k        = k_folds,
  nIter    = n_iter,
  CVMet    = cv_met,
  b_iter   = 5000, b_burnin = 1000, b_thin = 10,
  b_R2     = 0.5,  b_digits = 4,
  A_ext    = G_tuned   # NULL = use A.mat() internally
)
print(summarize_MT_Fits(cv_mt, CVMet = cv_met, traits = traits_mt))
```

### 2i — SE:MT GEBV Prediction

```{r se-mt-gebv}
gebv_mt <- getRankedPredictedValuesMT(
  pred_data,
  nTraits   = length(traits_mt),
  trait     = traits_mt,
  GPModelMT = "BRR (BGLR)",   # "RKHS (BGLR)", "Spike-Slab (BGLR)"
  optTS     = opt_ids,
  b_iter    = 5000, b_burnin = 1000, b_thin = 10,
  b_R2      = 0.5,  b_digits = 4,
  A_ext     = G_tuned
)
# gebv_mt[[1]] = ranked training predictions
# gebv_mt[[2]] = ranked target predictions

writeCVOutTable(cv_mt,        type = "MT", outDirPath = out_dir)
writeGPOutTable(gebv_mt[[1]], type = "MT", outDirPath = out_dir)
```

---

# Section 3 — ME Workflow: Multi-Environment

## 3a — Load ME Phenotype & Standardise Column Names

The ME phenotype CSV must have one row per observation (strain × environment).
`stdizePhVarNames()` renames user-supplied column names to the standard names
expected by downstream ME functions.

```{r me-pheno}
pheno_me_raw <- read.csv(me_pheno_path, stringsAsFactors = FALSE)
id_cols_me   <- colnames(pheno_me_raw)[!colnames(pheno_me_raw) %in% trait_me]

std_out      <- stdizePhVarNames(pheno_me_raw,
                  IDColsME   = id_cols_me,
                  StrainME   = strain_col,
                  EnvVarME   = env_col,
                  LocMEColVar = loc_col,
                  YrMEColVar  = yr_col)
pheno_me_mod <- std_out[[1]]   # standardised phenotype data frame
id_cols_mod  <- std_out[[2]]   # updated ID column vector
env_var_mod  <- std_out[[3]]   # standardised environment variable name

## Format into per-trait data frames required by getMergedDataME
pheno_me_fmt <- getPhenoMEData(PhenoME    = pheno_me_mod,
                               TraitME    = trait_me,
                               nSelTraits = length(trait_me),
                               IDColsME   = id_cols_mod,
                               StrainME   = strain_col)
```

## 3b — Merge ME Genotype + Phenotype

```{r me-merge}
merged_me <- getMergedDataME(phData    = pheno_me_fmt,
                              genoImp   = geno_imp,
                              TargetIDs = NULL)
DT_list   <- merged_me[[1]]   # list of phenotype data frames, one per trait
geno_list <- merged_me[[2]]   # list of genotype matrices, one per trait
```

## 3c — (Optional) Enviromics: Build Environmental Kernel (KE)

`KE` is required for `FitEnvModels = TRUE` in ME CV and prediction.
Skip this block and set `KE <- NULL` if no environmental covariate data
is available.

```{r enviromics}
# coords_df <- read.csv("<path/to/location_coordinates.csv>")
# # Expected columns: Location, Country, Lat, Long
#
# env_data  <- getEnvData(coords_df,
#                startDate = "2020-04-01",
#                endDate   = "2020-10-31")
#
# env_ker   <- getEnvKernel(env_data, process = TRUE, Gaussian = FALSE)
# env_ker   <- syncEnvPhenoDat(env_ker,
#                LocCoord = coords_df,
#                OtherLoc = FALSE)
# KE        <- env_ker$W          # symmetric env × env kernel
# plotEnvRel(env_ker, outDirPath = out_dir)

KE <- NULL   # replace with env_ker$W when enviromics data is available
```

## 3d — ME Cross-Validation

Multi-environment CV using **BGGE** kernels.
`KG = G_tuned` supplies the genomic kinship matrix; `KG = NULL` estimates it
internally. `FitEnvModels = !is.null(KE)` automatically enables/disables the
environmental component.

```{r me-cv}
cv_me <- getME_CV(
  DT_1_Filt_List = DT_list,
  genoDat_List   = geno_list,
  traits         = trait_me,
  KG             = G_tuned,        # NULL = estimated internally
  KE             = KE,
  CVMet          = cv_met,
  factVar        = env_var_mod,
  K              = k_folds,
  NIter          = n_iter,
  KMethod        = "Linear",       # "Linear" or "Gaussian"
  FitEnvModels   = !is.null(KE),
  fixedME        = NULL,
  envVar         = env_var_mod,
  IDColsME       = id_cols_mod,
  IDColME        = env_var_mod
)
cv_me_summary <- summarize_ME_CV_fits(cv_me, cvMet = cv_met)
print(cv_me_summary)
writeCVOutTable(cv_me_summary, type = "ME", outDirPath = out_dir)
```

## 3e — ME GEBV Prediction (Leave-One-Factor-Out)

LOFO prediction masks one environment at a time.
`maskFactrLev` specifies the environment(s) to mask; `maskProp = 1` masks all
genotypes in that level.

```{r me-gebv}
me_pred <- fitMEModels_LOFO_Pred(
  DT_1_Filt_List = DT_list,
  genoDat_List   = geno_list,
  traits         = trait_me,
  KG             = G_tuned,
  KE             = KE,
  factrVar       = env_var_mod,
  maskFactrLev   = unique(pheno_me_mod[[env_var_mod]])[1],
  maskProp       = 1,
  nSampleReps    = 1,
  fixedME        = NULL,
  envVar         = env_var_mod,
  IDColsME       = id_cols_mod,
  IDColME        = env_var_mod,
  KMethod        = "Linear",
  FitEnvModels   = !is.null(KE)
)
```

## 3f — Combine and Write ME Output

```{r me-output}
me_combined <- getCombinedTab(
  outputListME = me_pred,
  TraitME      = trait_me,
  IDColsME     = id_cols_mod,
  IDColME      = env_var_mod,
  envVar       = env_var_mod,
  fitEnvCov    = !is.null(KE),
  reaction     = FALSE
)
writeGPOutTable(me_combined, type = "ME", outDirPath = out_dir)
cat("Pipeline complete. Results written to:", out_dir, "\n")
```

---

# Accessing the Raw Script

The bare R script (no prose, all executable code) is installed alongside the
package and can be opened directly:

```{r open-script}
file.edit(system.file("pipeline", "GS4PB_Pipeline.R", package = "GS4PB"))
```
